Large-scale spatiotemporal variation in vital rates and population dynamics of an alpine bird

### 2024-01-11 Manuscript_Nater_etal.qmd
Added draft text for "Limitations and outlook" section

Density dependence is an important process to understand in ecology but more so for exploited species like willow ptarmigan [@andrewartha1954; @sandercock_etal2011; @Aanes_Engen_SÆther_Willebrand_Marcström_2002; @willebrand2001]. Negative density dependence has been found in numerous gallinaceous birds such as northern bobwhites [*Colinus virginianus*; [@mcconnell2018]\], \[*Perdix perdix*; [@bro2003]\], and wild turkeys \[*Meleagris gallopavo*; [@mcghee2007]]*.* <!--# Got the tasty ones in here for you -->While we found the highest population growth rates in areas with relatively low density, and relatively low growth rates in high-density areas [@fig-paramCorrs A], our analysis did not provide evidence for a strong relationship between density and growth rates across space. However, this is unsurprising as we did find substantial spatial differences in demography, which suggests that carrying capacity too varies over space. Also, support for density dependence in willow ptarmigan has been mixed. @myrberget1988 observed no change in productivity despite a 50% decrease in abundance. Conversely, @pedersen2004 found strong overwinter negative density dependence and they posited that immigration was the vital rate that responded to changes in density. Low survival rates appeared to have a one-year lag following abundance peaks. This pattern could be a result of negative density dependence that went undetected when looking at the ensemble of samples across the 40 areas. Negative density dependence has been shown in willow ptarmigan populations which may explain explain reduced survival following years with elevated densities [@pedersen2004].

<!--# Not happy with the DD para yet but I can work on Jan 11 -->

@@

-   Some limitations of "base model" that need mentioning (e.g. age assignment, ignoring potential effects of flock size on detection)

-   Outlook: improved modelling of spatio-temporal patterns through correlated REs

Another approach to maximising information gain when jointly analysing data from different areas is explicit modelling of spatial and spatio-temporal variation. While this is commonly done for modern species distribution models [e.g. @pacifici2017; @guelat2018], it has rarely been implemented in demographic models. This is not least due to the fact that few demographic models span large spatial extents. Since our IDSM and the willow ptarmigan data do actually do that though, extending the framework to account for spatial and spatiotemporal correlations may be feasible. Per now, we have not yet attempted this, but are planning to do so in the near future. (BUT: we do already have some computational limitations now, so likely to need development of more efficient sampling strategies. Which is fortunately possible with NIMBLE but there is a chance that another platform, e.g. INLA, may be required) <!-- @ERLEND: related to this, I want to discuss the potential of including the not publicly available data. Is there an overview somewhere over what the coverage (in terms of municipalities) would be if we had access to all the Hønsefuglportalen data? This might matter not just for increasing sample size, but perhaps also making CAR modelling (spatial structure through neighborhood relations) possible -->

<!-- @MYSELF: Have a closer look at chapter 19 in IPM book to see what they did there and also what other work they reference -->

-   Limited usefulness of auxiliary radio-telemetry data as only from one area (big potential for including additional datasets from other small-scale studies, telemetry but also other, e.g. nest monitoring?) <!-- @ERLEND: Any thoughts on what the status here is? Might there be other data available from beyond Lierne? Are there exciting future prospects e.g. linked to automated monitoring with cameras/sound boxes at a larger scale? -->

    The framework used in our study can be adapted to include the effects of management actions and harvest. Harvest effects on willow ptarmigan has been studied yet uncertainties still exist [@sandercock_etal2011; @Aanes_Engen_SÆther_Willebrand_Marcström_2002; @pedersen2004; @willebrand2001]. Also, there is a dearth of empirical information about the effects of harvest on many exploited species despite decades of continued exploitation.

-   <!-- @JAMES: This one is for you. Some ruminations on how and why it would be great to formally model harvest, integration with hunting data, maybe eventually develop a tool to calculate quotas (I know this has been developed for other game species), etc. -->

-   <!-- Will add more Jan 11 -->

@@

-   Also: useful to link up to different reporting routes and/or more general (multi-indicator) projects <!-- @EVERYONE: Do we know of some good references for this? Prseumably we cannot cite our chapter in the Technical Manual yet :-P -->

@@

TBA ... We are grateful to the participants in the "Hønsefuglportalen" and their furry friends that collected ptarmigan observations. Additionally, we would like to thank Bernardo Brandão Niebuhr dos Santos for his help with spatial data visualization. J. A. Martin was supported by a Fulbright Fellowship and the Norwegian Institute for Nature Research that facilitated this collaboration. <!-- @ERLEND: you maybe know best all the people/institutions we need to thank here -->

>>>>

Density dependence is an important process to understand in ecology but more so for exploited species like willow ptarmigan [@andrewartha1954; @sandercock_etal2011; aanes2002; @willebrand2001]. Negative density dependence has been found in numerous gallinaceous birds such as northern bobwhites *Colinus virginianus* [@mcconnell2018], *Perdix perdix* [@bro2003], and wild turkeys *Meleagris gallopavo* [@mcghee2007]).

While we found the highest population growth rates in areas with relatively low density, and relatively low growth rates in high-density areas [@fig-paramCorrs A], our analysis did not provide evidence for a strong relationship between density and growth rates across space. However, this is unsurprising as we did find substantial spatial differences in demography, which suggests that carrying capacity too varies over space. Also, support for density dependence in willow ptarmigan has been mixed. @myrberget1988 observed no change in productivity despite a 50% decrease in abundance. Conversely, @pedersen2004 found strong overwinter negative density dependence and they posited that immigration was the vital rate that responded to changes in density. Low survival rates appeared to have a one-year lag following abundance peaks. This pattern could be a result of negative density dependence that went undetected when looking at the ensemble of samples across the 40 areas. Negative density dependence has been shown in willow ptarmigan populations which may explain explain reduced survival following years with elevated densities [@pedersen2004].

<!--# Not happy with the DD para yet but I can work on Jan 11 -->
<!--# @JAMES: Feel free to add more, but I feel like if we want to go for a DD deep-dive, I need to run a test for evidence of WITHIN area DD first. Because there may well be a pattern there... -->

@@

The primary focus of this work was placed on developing an effective pipeline for integrating data and modelling population dynamics across a large number of areas. Consequently, many additional opportunities for improving and refining the modelling framework itself remain. 
First, the precision and accuracy of model estimates might be increased through better accounting for heterogeneity and potential biases in detection of birds during the line transect surveys. In an earlier study analyzing data from the same monitoring programme, @bowler2020, for example, found that detection probability was not independent of the size of group birds were part of, resulting in birds in larger groups being more likely to be detected, especially at larger distances. When birds are observed in larger groups, it is also not unlikely that human observers may miscount, i.e. that there is some observation error in the number reported. This could be accounted for by adding an additional layer of hierarchy to observation process (see e.g. REF), which could further be used to also account for error in judging the observation distance [e.g. @marques2004]. Another potential source of bias in our IDSM is related to failure to correctly assign the age class of observed birds. @nilsen2023 showed that incorrect age assessment can bias (relative) estimates of survival and recruitment, and while they only found a weak bias in their case study on a single area, the problem may be larger in a multi-area setting that may contain areas with different proportions of misclassified observations. If misclassification happened at random, mixture models could be used to determine the likely age class of individuals to whom no age class was assigned during observations (+ref). In our case, however, juveniles are likely more prone to misclassification due to [...] <!-- @ERLEND can you add reasons? -->. Future studies should investigate to what degree available information on e.g. group composition could be used for this, and what kind of auxiliary data would need to be collected to reliably model misclassification error.

The second (and perhaps most attractive) aspect of our modelling framework in the context of future work is its spatio-temporal hierarchical structure. While we included spatial, temporal, and residual variation in our framework here, we treated them as independent. Alternatively, spatial (and temporal) correlations among parameters can be modelled explicitly, something that is commonly done e.g. for modern species distribution models [e.g. @pacifici2017; @guelat2018]. For demographic models, this has rarely been implemented so far, not least due to the fact that few demographic models have sufficient spatial resolution [@schaub2021]. The ptarmigan IDSM presented in this study, however, does have sufficient resolution and our results do indeed support that there is spatial clustering in both overall and time-dependent demographic parameters (e.g. @fig-pVR, SFs "Avg_pSurv_Map.pdf" & "Avg_rRep_Map.pdf"). Furthermore, we did find that mixing of several of the global and area-specific intercept parameters in the current model was suboptimal, suggesting that there may be much to gain from additional structuring, as well as from development of more efficient MCMC sampling strategies for the resulting extended model. One promising framework for approaching this are conditionally autoregressive models (CARs). Such models have been used repeatedly for modelling spatial autocorrelation in species occupancy and demographic rates [e.g. @saracco2010; @saracco2012; @guelat2018]
and are straightforward to implement using NIMBLE [@lawson2020]. One possible challenge with using CAR models to explicitly model spatial correlations within our ptarmigan IDSM is that CAR models rely on "neighborhood" relationships between discrete areas and many "neighbors" are missing in our ptarmigan data (e.g. @fig-dens). While estimation of latent parameters in missing areas may possible (Perry de Valpine, personal communication; @schaub2021 chapter 19), this also provides an opportunity for inclusion of additional data. The line transect survey data included in this study constitutes just the publicly available part of the data collected through "Hønsefuglportalen" but the programme also includes additional surveys on private land. [...] <!-- @ERLEND: Can we add a very brief ocverview over what the coverage (in terms of municipalities) would be if we had access to all the Hønsefuglportalen data? --> Exploring to what degree additional data from Hønsefuglportalen could be included in future studies employing an extended IDSM with additional spatial structuring is therefore a worthwhile endeavor.

Finally, including further data beyond the line transect surveys may be relevant in the future, and in particular in the context of informing and improving management of ptarmigan hunting. 
In the present study, we have used auxiliary radio-telemetry data to supplement information on survival, but since this data was available for only one out of 41 areas, its influence was likely small. Nonetheless, this illustrates a way for how smaller datasets from single or subsets of areas can be integrated into a large-scale modelling framework. Other relevant data could be included using the same approach [...]  <!-- @ERLEND: Any thoughts on what the status here is? Might there be other data available from Lierne and/or beyond (other telemetry data, nest monitoring, etc.? Are there exciting future prospects e.g. linked to automated monitoring with cameras/sound boxes at a larger scale? -->
The most relevant source of data to be included into the IDSM framework in the near future, however, is harvest data. <!-- @ERLEND 1-2 sentence summary of what kind of harvest data is available, and at what spatial covarage and resolution --> As the IDSM framework is, in essence, an IPM, harvest can be modelled through partitioning of survival into cause-specific mortality in the process model and inclusion of relevant harvest data likelihoods [e.g. @gamelon2021; @nater2021]. Harvest effects on willow ptarmigan have been studied previously yet uncertainties persist [@sandercock_etal2011; aanes2002; @pedersen2004; @willebrand2001]. [...] <!-- @JAMES & Erlend: Maybe we can be a bit more specific about what the "uncertainties" are regarding ptarmigan harvest? Might be good to keep the "spatially explicit" perspective a bit, i.e. why would it be great to look at this with this kind of model, instead of just aggregate numbers across large regions? Also, maybe we should highlight that up-to-date (easilty repeatable) analyses are important because things can change, and harvest effects may interact with other (emergent) factors such as climate change and habitat degradation. Finally, it could also be nice to link this to some studies on other species... -->

@@

-   Also: useful to link up to different reporting routes and/or more general (multi-indicator) projects <!-- @EVERYONE: Do we know of some good references for this? Presumably we cannot cite our chapter in the Technical Manual yet :-P -->

@@

TBA ... We are grateful to the participants in the "Hønsefuglportalen" and their dog helpers that collected ptarmigan observations. Additionally, we would like to thank Bernardo Brandão Niebuhr dos Santos for his help with spatial data visualization. J. A. Martin was supported by a Fulbright Fellowship and the Norwegian Institute for Nature Research that facilitated this collaboration. <!-- @ERLEND: you maybe know best all the people/institutions we need to thank here -->
